cardinalR: Generating interesting high-dimensional data structures

### 2025-11-07 paper-cardinalR.Rmd
Fixed the latex issue

The simplest structures are approximately linear branches in \gD{}, generated by the `gen_linearbranches(n, k)` function. These consist of $k$ short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch $i$ is defined as

@@

To introduce curvature, the `gen_curvybranches(n, k)` function generates $k$ curvilinear branches in \gD{}. Branches $1$ and $2$ are simple parabolas defined as

@@

The `gen_expbranches(n, k)` function creates $k$ exponential branches in \gD{}, radiating from a central region. Each branch $i$ is defined as

@@

The function `gen_gridcube(n, p)` is a wrapper around `geozoo::cube.solid.grid()`. It generates a regular lattice of points in \pD{}, producing a uniform hypercube grid. Each axis contains equally spaced coordinates, resulting in a well-defined geometric structure.

@@

This structure appears in \pD{} data when variation is driven by a single factor, such as time-course or sensor measurements, providing a useful test case for trajectory and regression methods [@trapnell2014].

@@

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in \gD{} space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:polynomial-tb-html)', '\\@ref(tab:polynomial-tb-pdf)'))` summarizes these functions and their purposes.

@@

A pyramid structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr)"} else { "Figure \\@ref(fig:pyr-proj)"})`) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in \pD{} space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:pyramid-tb-html)', '\\@ref(tab:pyramid-tb-pdf)'))` summarizes these functions.

@@

The Swiss roll is a standard nonlinear manifold, representing a \gD{} plane curled into $3\text{-}D$. The `gen_swissroll(n, w)` generates points as $X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \quad t \sim U(0, 3\pi).$

>>>>

The simplest structures are approximately linear branches in $2\text{-}D$, generated by the `gen_linearbranches(n, k)` function. These consist of $k$ short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch $i$ is defined as

@@

To introduce curvature, the `gen_curvybranches(n, k)` function generates $k$ curvilinear branches in $2\text{-}D$. Branches $1$ and $2$ are simple parabolas defined as

@@

The `gen_expbranches(n, k)` function creates $k$ exponential branches in $2\text{-}D$, radiating from a central region. Each branch $i$ is defined as

@@

The function `gen_gridcube(n, p)` is a wrapper around `geozoo::cube.solid.grid()`. It generates a regular lattice of points in $p\text{-}D$, producing a uniform hypercube grid. Each axis contains equally spaced coordinates, resulting in a well-defined geometric structure.

@@

This structure appears in $p\text{-}D$ data when variation is driven by a single factor, such as time-course or sensor measurements, providing a useful test case for trajectory and regression methods [@trapnell2014].

@@

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in  space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:polynomial-tb-html)', '\\@ref(tab:polynomial-tb-pdf)'))` summarizes these functions and their purposes.

@@

A pyramid structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr)"} else { "Figure \\@ref(fig:pyr-proj)"})`) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in $p\text{-}D$ space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:pyramid-tb-html)', '\\@ref(tab:pyramid-tb-pdf)'))` summarizes these functions.

@@

The Swiss roll is a standard nonlinear manifold, representing a  plane curled into $3\text{-}D$. The `gen_swissroll(n, w)` generates points as $X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \quad t \sim U(0, 3\pi).$

### 2025-11-07 paper-cardinalR.tex
Fixed the latex issue

The simplest structures are approximately linear branches in $2\text{-}D${}, generated by the \texttt{gen\_linearbranches(n,\ k)} function. These consist of \(k\) short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch \(i\) is defined as

@@

To introduce curvature, the \texttt{gen\_curvybranches(n,\ k)} function generates \(k\) curvilinear branches in $2\text{-}D${}. Branches \(1\) and \(2\) are simple parabolas defined as

@@

The \texttt{gen\_expbranches(n,\ k)} function creates \(k\) exponential branches in $2\text{-}D${}, radiating from a central region. Each branch \(i\) is defined as

@@

The function \texttt{gen\_gridcube(n,\ p)} is a wrapper around \texttt{geozoo::cube.solid.grid()}. It generates a regular lattice of points in $p\text{-}D${}, producing a uniform hypercube grid. Each axis contains equally spaced coordinates, resulting in a well-defined geometric structure.

@@

This structure appears in $p\text{-}D${} data when variation is driven by a single factor, such as time-course or sensor measurements, providing a useful test case for trajectory and regression methods \citep{trapnell2014}.

@@

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in $2\text{-}D${} space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table \ref{tab:polynomial-tb-pdf} summarizes these functions and their purposes.

@@

A pyramid structure (Figure \ref{fig:pyr-proj}) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in $p\text{-}D${} space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions. Table \ref{tab:pyramid-tb-pdf} summarizes these functions.

@@

The Swiss roll is a standard nonlinear manifold, representing a $2\text{-}D${} plane curled into \(3\text{-}D\). The \texttt{gen\_swissroll(n,\ w)} generates points as \(X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \quad t \sim U(0, 3\pi).\)

>>>>

The simplest structures are approximately linear branches in \(2\text{-}D\), generated by the \texttt{gen\_linearbranches(n,\ k)} function. These consist of \(k\) short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch \(i\) is defined as

@@

To introduce curvature, the \texttt{gen\_curvybranches(n,\ k)} function generates \(k\) curvilinear branches in \(2\text{-}D\). Branches \(1\) and \(2\) are simple parabolas defined as

@@

The \texttt{gen\_expbranches(n,\ k)} function creates \(k\) exponential branches in \(2\text{-}D\), radiating from a central region. Each branch \(i\) is defined as

@@

The function \texttt{gen\_gridcube(n,\ p)} is a wrapper around \texttt{geozoo::cube.solid.grid()}. It generates a regular lattice of points in \(p\text{-}D\), producing a uniform hypercube grid. Each axis contains equally spaced coordinates, resulting in a well-defined geometric structure.

@@

This structure appears in \(p\text{-}D\) data when variation is driven by a single factor, such as time-course or sensor measurements, providing a useful test case for trajectory and regression methods \citep{trapnell2014}.

@@

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table \ref{tab:polynomial-tb-pdf} summarizes these functions and their purposes.

@@

A pyramid structure (Figure \ref{fig:pyr-proj}) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in \(p\text{-}D\) space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions. Table \ref{tab:pyramid-tb-pdf} summarizes these functions.

@@

The Swiss roll is a standard nonlinear manifold, representing a plane curled into \(3\text{-}D\). The \texttt{gen\_swissroll(n,\ w)} generates points as \(X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \quad t \sim U(0, 3\pi).\)
